Modeling mental rotation

### 2014-01-29 man/cogsci-2014/cogsci-2014.tex
More work on introduction/background/modeling

Previous models of mental rotation have, for the most part, not
addressed these questions. \citeA{Kosslyn:1977tv} focus primarily on
simulating the visual simulation buffer rather than the decision of
\textit{what} should be simulated. Similarly, \citeA{Glasgow:1992tj}
are more concerned with the representation of imagery than the
process. Although \citeA{Anderson1978} emphasizes the importance of
considering both representation and process, he dismisses the problem
of determining the direction and axis of rotation as merely a
``technical difficulty''.

The only model (of which the authors are aware) that has more
seriously attempted to address these concerns is the parallel-process
model of mental rotation by \citeA{Funt:1983wn}. He determines the
direction and axis of rotation using the method-of-moments method;
once this target rotation computed, his model rotates one object until
the target rotation is reached, and then compares the two to see if
they are congruent.

More recently, evidence has come to light that people \textit{do not}
actually rotate until the objects are congruent. Eye-tracking studies
suggest people pay attention to certain salient features of the
object, rather than a holistic representation \cite{Just1976}, and 
\citeA{Yuille:1982tx} e.g. rotating away from near-perfect matches \cite{Gardony:2013gn}

@@

Formally, we denote the shapes as $X_a$ and $X_b$ and assume $X_b$ is
generated by a transformation of $X_a$, i.e. $X_b=f(X_a, \theta, h)$,
where $\theta$ is a rotation, $h=0$ is the hypothesis that the images
depict the same object, and $h=1$ is the hypothesis that the images
depict mirror-image objects. The posterior probability of each
hypothesis given the observed shapes is then:
\begin{equation}
  p(h\ \vert\ X_a, X_b) \propto \int p(X_b\ \vert\ X_a, \theta, h)p(X_a)p(h)p(\theta)\dif\theta.
  \label{eq:posterior}
\end{equation}
Because we ultimately want to determine which hypothesis is more
likely, the quantity of interest is a ratio $\ell:=p(h=0\ \vert\ X_a,
X_b) / p(h=1\ \vert\ X_a, X_b)$ which (assuming the two hypotheses are
equally likely \textit{a priori}, and that all rotations are equally
likely) simplifies to:
\begin{equation}
  \ell = \frac{\int p(X_b\ \vert\ X_a, \theta, \hi)\dif\theta}{\int p(X_b\ \vert\ X_a, \theta, \hf)\dif\theta}.
  \label{eq:lh-ratio}
\end{equation}
If $\ell > 1$, then we accept the hypothesis that the images depict
the same object ($\hi$); if $\ell < 1$, then we accept the hypothesis
that the images depict flipped objects ($\hf$).

@@

If we represent a shape of $N$ vertices with a $N\times 2$ coordinate
matrix $\mathbf{X}=[\mathbf{x}_1, \ldots{}, \mathbf{x}_N]$, then the
transformation $f$ is $f(\mathbf{X}, h,
\theta):=\mathbf{X}\F_h^T\R^T$, where $\R$ is a rotation matrix,
$\F_0$ is the identity matrix $\I$, and $\F_1$ is a reflection matrix
across the $y$-axis. Assuming no computational constraints, the
simplest solution is to compute the left inverse of $\Xa\F_h^T$ and,
for each $h$, check whether $(\Xa
\F_h^T)_\mathrm{left}^{-1}\cdot{}\Xb$ is a valid rotation matrix.

However, we assume that the ability to compute left inverses is not
available. Instead, the observed shapes may only be transformed by a
small amount at a time, and each transformation is costly in terms of
computational resources. The goal, then is to estimate the integrals
in Eq.~\ref{eq:lh-ratio} by evaluating $p(\Xb\ \vert\ \Xa, \theta, h)$
as few times as possible.

@@

@@

The previous hypothesis in the literature is that people perform a
mental rotation in the shortest direction until a match is found, and
that determining this direction is an implementation detail \TODO{cite
  anderson?}. To reflect this hypothesis, we created an ``oracle''
model which knows the correct rotation \textit{a priori}, and simply
performs rotations in the shortest direction until it reaches the
point where the shapes are aligned.

@@

An intuitive and simple model which does not assume the direction is
known might first rotate in the direction that increases similarity,
and then stop rotating once a ``match'' is found, as determined by a
threshold on the value of $S$. If all rotations have been exhausted,
then flip, and try rotating again.

@@

Unfortunately, it is not always clear what constitutes a good enough
``match''. In the current formulation of the problem, we know the
exact geometry of the shapes, and know that a linear transformation
exists which will perfectly align the shapes. However, more generally,
shapes could be more complex and/or three-dimensional, be subject to
perceptual uncertainty, or even be not exactly identical. In these
cases, the problem of choosing a threshold becomes much harder: it is
not clear how to choose a reasonable threshold \textit{a priori}, as
the global optimum will depend on many factors. Indeed, shape matching
is a non-trivial problem in computer vision, requiring sophisticated
algorithms to achieve reasonable performance
\cite<e.g.,>{Belongie:2002tj,Sebastian:2003vm}.

Thus, the \Th{} model will not be robust beyond this domain. One
obvious alternative would be to perform an exhaustive search or use a
global optimisation strategy. However, this would not result in the
linear response time found by \citeA{Shepard1971}. A second
alternative (which would potentially demonstrate the linear response
time result) could instead just perform a Hill Climbing (\Hc{}) search
for a \textit{local} maximum. Similar to the \Th{} model, we use the
(local) MLE estimate of $\theta$, as in Eq.~\ref{eq:mle-lh-ratio}.

@@

An alternative to the previous models is not to perform a search for
the MLE, but to approximate the integrals in Eq.~\ref{eq:lh-ratio} by
picking locations to sample which will most improve accuracy of the
estimate. \TODO{Really need to motivate this better: why do we need
  something so complex? Talk about how it's robust to the shape of the
  function, the scale, noise, etc. It's complicated, but it should be
  able to handle these things, so let's try it and see if it works at
  all. If it does, we can work towards somethign more succinct in the
  future.}

@@

We ran model simulations for the same stimuli that participants
viewed. We ran 10 samples of each stimulus under the \Oc{} model, and
100 samples of each stimulus for the \Th{}, \Hc{}, and
\Bq{} models, with the exception of stimuli where
$\theta=0$ or $\theta=180$, in which case we ran double the number of
samples.

>>>>

In this paper, we attempt to address these questions through a
rational analysis \cite{Marr:1983to,anderson90,Shepard:1987tt} of four
models of mental rotation. First, we discuss the previous literature
on mental imagery. Second, we outline computational- and
algorithmic-level analyses of the problem of mental rotation. Then, we
describe a behavioral experiment based on the classic mental rotation
studies \cite<e.g.,>{Cooper:1975wp}, and compare the results of the
experiment with each of the models. We conclude with a discussion of
the strengths and weaknesses of each model, and lay out directions for
future work.

Previous models of mental rotation have, for the most part, focused
mostly on the representation of mental images, rather than how people
decide \textit{which} mental images to
simulate. \citeA{Kosslyn:1977tv} propose a model of mental imagery's
visual buffer, but not the planning process of how it is used;
similarly, \citeA{Glasgow:1992tj} are mostly concerned with the
knowledge representation of imagery. Although \citeA{Anderson1978}
emphasizes the importance of considering both representation and
process, he dismisses the problem of determining the direction and
axis of rotation as merely a ``technical difficulty''.

The only model (of which the authors are aware) that has more
seriously attempted to address concerns regarding which simulation to
run is the parallel-process model of mental rotation by
\citeA{Funt:1983wn}. He determines the direction and axis of rotation
using the method-of-moments method; once this target rotation
computed, his model rotates one object until the target rotation is
reached, and then compares the two to see if they are
congruent. \TODO{?}

However, there is evidence that people do not actually rotate until
the objects are congruent, but rather pay attention to certain salient
features of the object, rather than a holistic representation
\cite{Just1976,Yuille:1982tx}. More recent evidence shows that when
people perform physical rotations, they do not rotate until congruency
is reached and may even rotate \textit{away} from near perfect matches
\cite{Gardony:2013gn}. Moreover, the state-of-the-art in computer
vision suggests that there is more to shape matching than simply
checking for congruency, particularly when shapes are complex or when
the shapes might not be exactly the same
\cite<e.g.,>{Belongie:2002tj,Sebastian:2003vm}.

@@

Formally, we denote the shapes as $X_a$ and $X_b$ and assume $X_b$ is
generated by a transformation of $X_a$, i.e. $X_b=f(X_a, \theta, h)$,
where $\theta$ is a rotation, $h=0$ is the hypothesis that the images
depict the same object, and $h=1$ is the hypothesis that the images
depict mirror-image objects. The posterior probability of each
hypothesis given the observed shapes is then:
\begin{equation}
  p(h\ \vert\ X_a, X_b) \propto \int p(X_b\ \vert\ X_a, \theta, h)p(X_a)p(h)p(\theta)\dif\theta,
  \label{eq:posterior}
\end{equation}
where $p(X_b\ \vert\ X_a, \theta, h)$ is the probability that $X_b$
was generated from $X_a$, and $p(X_a)$ is the probability of observing
$X_a$. Because we ultimately want to determine which hypothesis is
more likely, the quantity of interest is a ratio $\ell:=p(h=0\ \vert\
X_a, X_b) / p(h=1\ \vert\ X_a, X_b)$ which (assuming that all
rotations are equally likely) simplifies to:
\begin{equation}
  \ell = \frac{\left(\int p(X_b\ \vert\ X_a, \theta, \hi)\dif\theta\right)p(h=0)}{\left(\int p(X_b\ \vert\ X_a, \theta, \hf)\dif\theta\right)p(h=1)}.
  \label{eq:lh-ratio}
\end{equation}
If $\ell > 1$, then we accept the hypothesis that the images depict
the same object ($\hi$); if $\ell < 1$, then we accept the hypothesis
that the images depict flipped objects ($\hf$).

@@

We represent a shape of $N$ vertices with a $N\times 2$ coordinate
matrix $\mathbf{X}=[\mathbf{x}_1, \ldots{}, \mathbf{x}_N]$, and denote
the rotation and/or reflection transformation as $f(\mathbf{X}, h,
\theta):=\mathbf{X}\F_h^T\R^T$, where $\R$ is a rotation matrix, and
$\F_h$ is either the identity matrix $\I$ (when $h=0$) or a reflection
matrix across the $y$-axis (when $h=1$).  We assume that the observed
shapes may only be transformed by a small amount at a time, and each
transformation is costly in terms of computational resources. The
goal, then is to estimate the integrals in Eq.~\ref{eq:lh-ratio} by
evaluating $p(\Xb\ \vert\ \Xa, \theta, h)$ as few times as possible.

@@

To summarize, we approximate the likelihood term of
Eq.~\ref{eq:lh-ratio} using the similarity function defined in
Eq.~\ref{eq:similarity}. Because we assume analog rotations, this
similarity can only be computed at sequential locations, i.e. the
actions listed in Eq.~\ref{eq:actions}.

\subsection{Specific models}

The last decision we need to make is which algorithm to use to
approximate Eq.~\ref{eq:lh-ratio} using samples of the similarity
function. The algorithm must do two things: first, it must choose
\textit{which} places to sample, and second, it must decide when to
stop sampling.

@@

One hypothesis in the literature is that people perform a mental
rotation in the shortest direction until a match is found
\cite{Shepard1971, Cooper:1975wp}, and that determining this direction
is an implementation detail \cite{Anderson1978}, or that it is
computed beforehand \cite{Funt:1983wn}.  To reflect this hypothesis,
we created an ``oracle'' model which first computes the correct
rotation and then simply takes actions in the shortest direction until
it reaches the point where the shapes are aligned.

To determine the correct rotation, we solve for the rotation matrix by
computing $(\Xa \F_h^T)_\mathrm{left}^{-1}\cdot{}\Xb$, where
$(\Xa\F_h^T)_\mathrm{left}^{-1}$ is the left inverse of
$\Xa\F_h^T$. For each $h$, we then check whether the resulting matrix
is a valid rotation matrix.

@@

An intuitive and simple model which does not know which way to rotate
\textit{a priori} might first rotate in the direction that increases
similarity, and then stop rotating once a ``match'' is found, as
determined by a threshold on the value of $S$. If all rotations have
been exhausted, then flip, and try rotating again.

We note, however, that it is not always clear what constitutes a good
enough ``match''. In the current formulation of the problem, we know
the exact geometry of the shapes, and know that a linear
transformation exists which will perfectly align the shapes. However,
shapes could be more complex and/or three-dimensional, be subject to
perceptual uncertainty, or even be not exactly identical. In these
cases, it is not clear how to choose a reasonable threshold \textit{a
  priori}, as the global optimum will depend on many factors. Thus, it
is unlikely that the \Th{} model will not be robust beyond this
domain.

@@

One way to deal with the problem of choosing a threshold would simply
be to perform an exhaustive search or use a global optimisation
strategy. However, these strategies would not result in the linear
response time found by \citeA{Shepard1971}. A second alternative,
which could potentially demonstrate the linear response time result,
might instead just perform a Hill Climbing (\Hc{}) search for a
\textit{local} maximum. However, it is not necessarily the case that
all local maxima for the correct hypothesis will be higher than all
the local maxima for the incorrect hypothesis: thus, we would expect
the \Hc{} model to have difficulties reliably choosing the correct
hypothesis.

@@

While the previous few models have all focused on \textit{searching}
for the global maximum, the goal is really to just approximate
Eq.~\ref{eq:lh-ratio} while still obeying the constraint of sequential
rotations. Instead of perfoming a search, we could try maintaining a
probability distribution over our \textit{estimate}, and then choose
actions which are expected to improve the estimate of
Eq.~\ref{eq:lh-ratio}. This strategy has several obvious
benefits. First, it does not make assumptions about the shape or scale
of the similarity function. Second, by choosing to sample places which
are informative, this method implicitly minimizes the amount of
rotation, but because it is constrained to sequential rotations,
should still exhibit the linear effect of the angle.

@@

We ran model simulations for the same stimuli that participants
viewed. We ran 10 samples of each stimulus under each of the modesl,
with the exception of stimuli where $\theta=0$ or $\theta=180$, in
which case we ran double the number of samples.
